An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-03 uai2017/paper.tex
Reorganized paper based on John's feedback.
Moved some preliminary stuff later, and added slightly more info. Also added a bit on hamiltonian dynamics. OK, still waiting for Xinlei and I to finish this.

%  \item Increase the minibatch size when needed. Log acceptance variance
%    scales as $1/k$ vs the minibatch size $k$. Our test is adaptive like earlier
%    works, but unlike them, the distribution of minibatch size is Gaussian, not
%    long-tailed.  Increased minibatch size also reduces the error rate for the
%    test.
\end{itemize}

@@

To address this challenge,~\citep{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis testing. Each
iteration, they start with a small minibatch and test whether $\theta'$ should
be accepted based on approximating $u<\alpha(\theta,\theta')$. If the
approximate test cannot decide with sufficient confidence, the minibatch grows
and the test repeats. This process continues until a decision. The bounds depend
on either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. We refer to these algorithms,
respectively, as {\sc AustereMH} and {\sc MHSubLhd}.

In~\citep{cutting_mh_2014}, CLT asymptotic arguments are used to show that their
approximate acceptance test error (with respect to full-batch M-H) tends to zero
as batch size increases, but no quantitative bounds are given. In addition, to
design a test that minimizes data usage given an error bound, they rely on
computing a standardized mean $\mu_{\rm std}$ each iteration. As the value of
$\mu_{\rm std}$ depends on all $\log p(x_i|\theta)$ and $\log p(x_i| \theta')$
terms, the test is technically not a minibatch test; \citep{cutting_mh_2014}
propose a conservative variant which assumes $\mu_{\rm std}=0$ and avoids the
need to evaluate all likelihood terms. We refer to this as {\sc AustereMH(c)}
and the default, non-conservative variant as {\sc AustereMH(nc)}.

For {\sc MHSubLhd}, the test requires direct bounds on the log likelihood ratio,
which for general distributions requires knowing $p(x_i | \theta)$ and $p(x_i |
\theta')$ for all $N$ samples. In~\citep{icml2014c1_bardenet14}, explicit bounds
are given and depend on
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i | \theta') - \log p(x_i | \theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression and other simple models, but it is difficult to do so for more
complicated models, so one generally needs to use all $p(x_i | \theta')$
terms\footnote{The sample code provided by~\citep{icml2014c1_bardenet14}
computes $C_{\theta,\theta'}$ by traversing the entire data, thus providing no
performance advantage over the complete test.} to get $C_{\theta,\theta'}$, like
the {\sc AustereMH(nc)} test.

In contrast, as we show in Section~\ref{sec:analysis}, we use quantitative forms
of the CLT which rely on measurable statistics from a \emph{single} minibatch.
In addition, while both {\sc AustereMH(nc)} and {\sc MHSubLhd} show useful
reductions in the number of samples required, they suffer the drawback of
resolving small log likelihood ratio differences between the minibatch and full
batch versions. We discuss a worst-case scenario in
Section~\ref{ssec:gaussian_example}.

@@

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, {\sc AustereMH(nc)} and
    {\sc MHSubLhd} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Appendix, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Lambda-\psi$
from zero; for certain input/random $u$ combinations, this difference can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Lambda^*$ to {\em replace} the variation due to $u$.

@@

We now derive error bounds for our M-H test and the target distribution it
generates. In Section~\ref{ssec:delta_star_distribution}, we present bounds on
the absolute and relative error (in terms of the CDFs) of the distribution of
$\Delta^*$ vs.  a Gaussian. We then show in Section~\ref{ssec:preserve_bounds}
that these bounds are preserved after the addition of other random variables
(e.g., $X_{\rm nc}$ and $X_{\rm corr}$). It then follows that the acceptance
test has the same error bound.

@@

\bibliography{paper}
%\bibliographystyle{plain}
\bibliographystyle{plainnat}

@@

For the Gaussian mixture experiment, we use the conservative method
from~\citep{cutting_mh_2014}, which avoids the need for recomputing log
likelihoods of each data point each iteration by choosing baseline minibatch
sizes $m$ and per-test thresholds $\epsilon$ beforehand, and then using those
values for the entirety of the trials. We experimented with the following
values:
\begin{itemize}[noitemsep]
\item $\epsilon \in \{0.001, 0.005, 0.01, 0.05, 0.1, 0.2\}$
\item $m \in \{50, 100, 150, 200, 250, 300, 350, 400, 450, 500\}$
\end{itemize}
The main difference between {\sc AustereMH(c)} and {\sc
AustereMH(nc)}\footnote{{\sc AustereMH(nc)} is used in
Section~\ref{ssec:logistic}.}  is that the latter needs to run a grid search
each iteration (i.e. after each time it makes an accept/reject decision for one
sample $\theta_t$). We use the same $\epsilon$ and $m$ candidates above for {\sc
AustereMH(nc)}.

>>>>

    \item Utilize Hamiltonian and Langevin Dynamics to judiciously model a system
    so as to generate distant yet high-quality proposals.
\end{itemize}

@@

To address this challenge,~\citep{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis testing. Each
iteration, they start with a small minibatch and test whether $\theta'$ should
be accepted based on approximating $u<\alpha(\theta,\theta')$. If the
approximate test cannot decide with sufficient confidence, the minibatch grows
and the test repeats. This process continues until a decision. The bounds depend
on either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. We refer to these algorithms,
respectively, as {\sc AustereMH} and {\sc MHSubLhd}. While both show useful
reductions in the number of samples required, they suffer from two drawbacks:
having to compute all log likelihood tokens each iteration to compute certain
statistics,\footnote{Thus, these are not technically minibatch methods.} (see
Section~\ref{sec:analysis}), and resolving small log likelihood ratio
differences between the minibatch and full batch versions. We discuss a
worst-case scenario in Section~\ref{ssec:gaussian_example}.

@@

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, {\sc AustereMH} and {\sc
    MHSubLhd} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Appendix, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Lambda-\psi$
from zero; for certain input/random $u$ combinations, this difference can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Lambda^*$ to {\em replace} the variation due to $u$.

@@

We now derive error bounds for our M-H test and the target distribution it
generates. In the most similar prior works,~\citep{cutting_mh_2014}, CLT
asymptotic arguments are used to show that their approximate acceptance test
error (with respect to full-batch M-H) tends to zero as batch size increases,
but no quantitative bounds are given. In addition, to design a test that
minimizes data usage given an error bound, they rely on computing a standardized
mean $\mu_{\rm std}$ each iteration. As the value of $\mu_{\rm std}$ depends on
all $\log p(x_i|\theta)$ and $\log p(x_i| \theta')$ terms, the test is
technically not a minibatch test; \citep{cutting_mh_2014} propose a conservative
variant which assumes $\mu_{\rm std}=0$ and avoids the need to evaluate all
likelihood terms. We refer to this as {\sc AustereMH(c)} and the default,
non-conservative variant as {\sc AustereMH(nc)}.

For {\sc MHSubLhd}, the test requires direct bounds on the log likelihood ratio,
which for general distributions requires knowing $p(x_i | \theta)$ and $p(x_i |
\theta')$ for all $N$ samples. In~\citep{icml2014c1_bardenet14}, explicit bounds
are given and depend on
\begin{equation}\label{eq:bad_bound}
    C_{\theta, \theta'} = \max_{1\leq i\leq N}|\log p(x_i | \theta') - \log p(x_i | \theta)|.
\end{equation}
It is possible to analytically determine $C_{\theta,\theta'}$ for logistic
regression and other simple models, but it is difficult to do so for more
complicated models, so one generally needs to use all $p(x_i | \theta')$
terms\footnote{The sample code provided by~\citep{icml2014c1_bardenet14}
computes $C_{\theta,\theta'}$ by traversing the entire data, thus providing no
performance advantage over the complete test.} to get $C_{\theta,\theta'}$, like
the {\sc AustereMH(nc)} test.  In contrast, we use quantitative forms of the CLT
which rely on measurable statistics from a \emph{single} minibatch.

In Section~\ref{ssec:delta_star_distribution}, we present bounds on the absolute
and relative error (in terms of the CDFs) of the distribution of $\Delta^*$
versus a Gaussian. We then show in Section~\ref{ssec:preserve_bounds} that these
bounds are preserved after the addition of other random variables (e.g., $X_{\rm
nc}$ and $X_{\rm corr}$). It then follows that the acceptance test has the same
error bound.

@@

\bibliography{paper}
\bibliographystyle{plainnat}

@@

For the Gaussian mixture experiment, we use the conservative method
from~\citep{cutting_mh_2014}, which avoids the need for recomputing log
likelihoods of each data point each iteration by choosing baseline minibatch
sizes $m$ and per-test thresholds $\epsilon$ beforehand, and then using those
values for the entirety of the trials. We experimented with the following
values:
\begin{itemize}[noitemsep]
\item $\epsilon \in \{0.001, 0.005, 0.01, 0.05, 0.1, 0.2\}$
\item $m \in \{50, 100, 150, 200, 250, 300, 350, 400, 450, 500\}$
\end{itemize}
and chose the $(m,\epsilon)$ pairing which resulted in the lowest expected data
usage given a selected upper bound on the error. Through personal communication
with~\citet{cutting_mh_2014}, we were able to use their same code to compute
expected data usage and errors.

The main difference between {\sc AustereMH(c)} and {\sc
AustereMH(nc)}\footnote{{\sc AustereMH(nc)} is used in
Section~\ref{ssec:logistic}.}  is that the latter needs to run a grid search
each iteration (i.e. after each time it makes an accept/reject decision for one
sample $\theta_t$). We use the same $\epsilon$ and $m$ candidates above for {\sc
AustereMH(nc)}.
